Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-10 paper/paper.tex
Changed some wording for precision preserving intro.

\documentclass[preprint,onecolumn,9pt]{sigplanconf} %{onecol}
\usepackage{alltt}
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{url}
\usepackage{graphicx}

@@

We discuss a variety of techniques for increasing performance by
decreasing the interpretive overhead of abstract interpretation, all
in the purely functional realm. These techniques are the following:
\begin{itemize}
 \item{shrink state space by quotients (lazy non-determinism)}
 \item{shrink state space by removing {\tt ev} states (abstract
   compilation)}
 \item{decrease intermediate data structures by specializing our
fixpoint operator to work with widened machine configurations
(fixpoint specialization)}
 \item{decrease intermediate data structures by operating on store
   changes rather than whole stores (store deltas)}
\end{itemize}
Each technique requires minimal changes to the reduction
relation/fixpoint, and each improves performance by integer factors.

@@

\subsection{Computing with store changes}

@@

% For David to fill in.
\subsection{Imperative global store and worklist}
...
\subsection{Pre-allocated global store}
...

@@

We define a relation between strict and lazy machines that illustrates
the quotient that the lazy states

\newcommand{\lapprx}[3]{#1 \approx_{#2} #3}
\newcommand{\apprx}[3]{#1 \sim_{#2} #3}
\newcommand{\capprx}[2]{#1 \sim #2}

\begin{align*}
\ev{\mexp, \menv, \mkont, \mcntr} \in cs &\Rightarrow \lapprx{\ev{\mexp, \menv,\mkont, \mcntr}}{\msto}{cs} \\
\co{\mkont, \mval} \in cs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \lapprx{\co{\mkont, \mval}}{\msto}{cs} \\
\ap{\mval_0, \mval_1, \mkont} \in cs \mbox{ and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \lapprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{cs} \\
\ans{\mval'} \in cs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \lapprx{\ans{\mval}}{\msto}{cs} \\
\ev{\mexp, \menv, \mkont} \in lcs &\Rightarrow \apprx{\ev{\mexp, \menv,\mkont}}{\msto}{lcs} \\
\co{\mkont, \mval} \in lcs \mbox { and } \mval' \in force(\msto, \mval) &\Rightarrow \apprx{\co{\mkont, \mval'}}{\msto}{lcs} \\
\ap{\mval_0, \mval_1, \mkont} \in lcs \mbox { and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \apprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{lcs} \\
\ans{\mval} \in lcs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \apprx{\ans{\mval'}}{\msto}{lcs} \\
\forall lc \in lcs, \lapprx{lc}{\msto}{cs}, \mbox{ and }\forall c \in cs, \apprx{c}{\msto}{lcs}. &\Rightarrow \capprx{(lcs, \msto)}{(cs, \msto)}
\end{align*}

@@

            \ev{\mexp_i, \menv, \mkont, \msto}$: \\

@@

      By cases on $\{\strue, \sfalse\} \cap force(\msto, \mval)$: \\
  \item{Case $\{\strue, \sfalse\}$:}
  \item{Case $\{\strue\}$:}
  \item{Case $\{\sfalse\}$:}
 ...}
%% \item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \mval, \msto} \machstep
%%             \ev{\mexp_1, \menv, \mkont, \msto}$: \\
%%       Where $\sfalse \in force(\msto, \mval)$ and $\mkont \in \msto(\maddr)$. \\
%%       ...}
\item{Case $\ap[^\mcntr_\mlab]{\mvalx{u}, \mval, \mkont,\msto} \machstep

@@

      {\bf BREAKDOWN}
      Where $\clos{\mvar, \mexp, \menv} \in force(\msto, \mvalx{u})$
        and $\menv', \msto', \mcntr' = \mathit{bind}^\mcntr_\mlab(\menv, \msto, \mvar, \mval)$
      ...}
\item{Case $\ap[^\mcntr_\mlab]{\mvalx{u}, \mval, \mkont,\msto} \machstep

@@

      {\bf BREAKDOWN}
      Where $\mop \in force(\msto, \mvalx{u})$, $\mval' \in force(\msto, \mval)$

@@

      ...}

@@

\begin{theorem}[Laziness preserves precision]
$\forall lcs,cs,lcs',\msto,\msto'$ if $\capprx{(lcs, \msto)}{(cs,\msto)}$
  and $(lcs,\msto) \machstep (lcs',\msto')$ then there exists $cs'$
  such that $(cs,\msto) \machstep (cs',\msto')$
\end{theorem}
\begin{proof}
Let $lc \in lcs$ be arbitrary and let $\hat{c} \subseteq cs$ be the
smallest set such that $\lapprx{lc}{\msto}{\hat{c}}$. \\
...
\end{proof}

>>>>

\documentclass[preprint,onecolumn,9pt]{sigplanconf} %{onecol}
\usepackage{alltt,mathpartir}
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{url}
\usepackage{graphicx}

@@

We discuss a variety of techniques for increasing performance by
decreasing the interpretive overhead of abstract interpretation. They
come in two flavors: decreasing the size of the represented state
space [SS], and decreasing intermediate data structures [DS].

\begin{itemize}
 \item{[SS] quotient by sets of values (lazy non-determinism)}
 \item{[SS] remove {\tt ev} states (abstract compilation)}
 \item{[DS] specialize the fixpoint operator to work with widened
   machine configurations (fixpoint specialization)}
 \item{[DS] operate on store changes rather than whole stores (store deltas)}
 \item{[DS] represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting (timestamp)}
 \item{[DS] use a global mutable workset}
 \item{[DS] represent the global store as a global, mutable vector in
   the 0-CFA case. (preallocation)}
\end{itemize}
Each technique requires minimal changes to the reduction
relation/fixpoint, and each improves performance by integer
factors. The first four remain in the purely functional realm. The
final three use mutation in small, localized places, so they do not
increase conceptual complexity drastically.

@@

\subsection{Store deltas}

@@

% For David to fill in.
\subsection{Timestamp}
...
\subsection{Pre-allocated global store}
...

@@

We define a relation between strict and lazy machines that illustrates
the quotient that the lazy states embody.

\newcommand{\vapprx}[3]{#1 \cong_{#2} #3}
\newcommand{\kapprx}[3]{#1 \cong_{#2} #3}
\newcommand{\lapprx}[3]{#1 \approx_{#2} #3}
\newcommand{\apprx}[3]{#1 \sim_{#2} #3}
\newcommand{\capprx}[2]{#1 \sim #2}

\begin{mathpar}
\inferrule{\mval' \in force(\msto, \mval)}
          {\vapprx{\mval}{\msto}{\mval'}} \quad
\inferrule{\ans{\mval} \in cs}{\lapprx{\ans{\mval}}{\msto}{cs}} \quad
\inferrule{\ev{\mexp, \menv, \mkont, \mcntr} \in cs}
          {\lapprx{\ev{\mexp, \menv,\mkont, \mcntr}}{\msto}{cs}} \\
\inferrule{\co{\mkont, \mval'} \in cs \\
           \vapprx{\mval}{\msto}{\mval'}}
          {\lapprx{\co{\mkont, \mval}}{\msto}{cs}} \quad
\inferrule{\ap{\mval_0, \mval_1, \mkont} \in cs \\
           \vapprx{\mval_1}{\msto}{\mval_1'}}
          {\lapprx{\ap{\mval_0, \mval_1', \mkont}}{\msto}{cs}} \\
\inferrule{\ev{\mexp, \menv, \mkont} \in lcs}
          {\apprx{\ev{\mexp, \menv,\mkont}}{\msto}{lcs}} \quad
\inferrule{\ans{\mval} \in lcs \\
           \vapprx{\mval}{\msto}{\mval'}}
          {\apprx{\ans{\mval'}}{\msto}{lcs}} \\
\inferrule{\co{\mkont, \mval} \in lcs \\
           \vapprx{\mval}{\msto}{\mval'}}
          {\apprx{\co{\mkont, \mval'}}{\msto}{lcs}} \quad
\inferrule{\ap{\mval_0, \mval_1, \mkont} \in lcs \\
           \vapprx{\mval_1}{\msto}{\mval_1'}}
          {\apprx{\ap{\mval_0, \mval_1', \mkont}}{\msto}{lcs}} \\
\inferrule{\forall lc \in lcs, \lapprx{lc}{\msto}{cs} \\
           \forall c \in cs, \apprx{c}{\msto}{lcs}}
          {\capprx{(lcs, \msto)}{(cs, \msto)}}
\end{mathpar}

@@

            \ev{\mexp_0, \menv, \mkont, \msto}$: \\

@@

      By definition of $\machstep$, $\strue \in force(\msto, \mval)$, thus by definition of
      $\capprx{}{}$, the strict machine takes the same step.}
            \ev{\mexp_1, \menv, \mkont, \msto}$: \\
      By definition of $\machstep$, $\sfalse \in force(\msto, \mval)$, thus by definition of
      $\capprx{}{}$, the strict machine takes the same step.}
\item{Case $\ap[^\mcntr_\mlab]{\clos{\mvar, \mexp, \menv}, \mval, \mkont,\msto} \machstep

@@

      Where $\menv', \msto', \mcntr' = \mathit{bind}^\mcntr_\mlab(\menv, \msto, \mvar, \mval)$
        $cs = \{\ap[^\mcntr_\mlab]{\clos{\mvar, \mexp, \menv}, \mval', \mkont} \mid
                \mval' \in force(\msto, \mval)\}$.
      Thus $cs' = \{\ev[^{\mcntr'}]{\mexp, \menv', \mkont}\}$ (clearly $\lapprx{lc'}{\msto'}{cs'}$)  and
      for all $c \in cs$, $wn(c,\msto) \machstep wn(c',\msto\sqcup[\mvar\mcntr' \mapsto \{c.\mval\}])$.
      Thus the union of these stores is equal to $\msto'$.}
\item{Case $\ap[^\mcntr_\mlab]{\mop, \mval, \mkont,\msto} \machstep

@@

      Where $\mval' \in force(\msto, \mval)$

@@

      By definition of $\capprx{}{}$ and $\interpdelta$,
        $cs = \{\ap[^\mcntr_\mlab]{\mop, \mvalx{u}, \mkont} \mid
                \mvalx{u} \in force(\msto, \mval), 
                \mval'' \in\interpdelta(\mop,\mvalx{u})\}$.
      Thus the strict machine takes the same step.}

@@

\begin{theorem}[Laziness preserves precision]
$\forall lcs,cs,lcs',\msto,\msto'$ if $\capprx{(lcs, \msto)}{(cs,\msto)}$
  and $(lcs,\msto) \machstep (lcs',\msto')$ then there exists $cs'$
  such that $(cs,\msto) \machstep (cs',\msto')$
\end{theorem}
\begin{proof}
By definition of $\machstep$,
$lcs' = \{lc' \mid \exists \msto^{lc}. wn(lc,\msto) \machstep wn(lc', \msto^{lc}), lc \in lcs\}$
For each $lc \in lcs$ let $\hat{c}^{lc} \subseteq cs$ be the
smallest set such that $\lapprx{lc}{\msto}{\hat{c}^{lc}}$. \\ By the
previous lemma, there exists a $\hat{c}^{lc}{}'$ such that $\capprx{(\{lc' \mid wn(lc,\msto) \machstep
  wn(lc', \msto^{lc})\}, \msto^{lc})}{(\hat{c}',\msto^{lc})}$. Let
$cs' = \bigcup\limits_{lc \in lcs}{\hat{c}^{lc}{}'}$ and $\msto' = \bigcup\limits_{lc \in lcs}{\msto^{lc}}$.
Since each step individually is related, the union is related: $\capprx{lcs'}{\msto'}{cs'}$.
By definition of $\machstep$, $(cs, \msto) \machstep (cs', \msto')$ and $(lcs, \msto) \machstep (lcs', \msto')$.
\end{proof}
